Optimizing Abstract Abstract Machines

### 2013-06-27 paper/paper.tex
Started cleaning up sentence breaks for better diffs.

\begin{abstract}
The technique of \emph{abstracting abstract machines} (AAM) provides a
systematic approach for deriving computable approximations of
evaluators that are easily proved sound.
%
This article contributes a complementary step-by-step process for
subsequently going from a \naive{} analyzer derived under the AAM
approach, to an efficient and correct implementation.  The end result
of the process is a two to three order-of-magnitude improvement over
the systematically derived analyzer, making it competitive with
hand-optimized implementations that compute fundamentally less precise
results.
\end{abstract}

@@

Program analysis provides sound predictive models of program behavior, but
in order for such models to be effective, they must be efficiently
computable and correct.  Past approaches to designing program analysis
have often featured abstractions that are far removed from the
original language semantics, requiring ingenuity in their construction
and effort in their verification.
%
The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides an alternative: a systematic way
of transforming a programming language semantics in the form of an
abstract machine into a family of abstract interpreters.  It thus reduces
the burden of constructing and verifying the soundness of an abstract
interpreter.
%

@@

By taking a
machine-oriented view of computation, AAM makes it possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Coq~\cite{local:harvard},
Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.

The primary strength of the approach is that abstract interpreters can
be easily derived through a small number of steps from existing
machine models.  Since the relationships between abstract machines and
higher-level semantic models---such as definitional
interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
these high-level semantic models to sound program analyzers in a
systematic way.  Moreover, since these analyses so closely resemble a
language's interpreter (a) implementing an analysis requires little
more than implementing an interpreter, (b) a single implementation can
serve as both an interpreter and analyzer, and (c) verifying the
correctness of the implementation is straightforward.

Unfortunately, the AAM approach yields analyzers with poor performance
relative to hand-optimized analyzers.
%
Our work takes aim squarely at this ``efficiency gap,'' and narrows it
in an equally systematic way through a number of simple steps, many of
which are inspired by run-time implementation techniques such as
laziness and compilation to avoid interpretative overhead.  Each of
these steps is proven correct, so the end result is an implementation
that is trustworthy and efficient.

In this article, we develop a systematic approach to deriving a
practical implementation of an abstract-machine-based analyzer using
mostly semantic means rather than tricky engineering. Our goal is to
empower programming language implementers and researchers to explore
and convincingly exhibit their ideas with a low barrier to entry. The
optimizations we describe are widely applicable and apparently
effective to scale far beyond the size of programs typically considered
in the recent literature on flow analysis for functional languages.

@@

\begin{figure}[t]
\small
\begin{center}
%\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of the rate of state
  transitions and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

We start with a quick review of the AAM approach to develop an
analysis framework and then apply our step-by-step optimization
techniques in the simplified setting of a core functional language.
This allows us to explicate the optimizations with a minimal amount of
inessential technical overhead.  Following that, we scale this
approach up to an analyzer for a realistic untyped, higher-order
imperative language with a number of interesting features and then
measure improvements across a suite of benchmarks.

At each step during the initial presentation and development, we
evaluated the implementation on a set of benchmarks. The highlighted
benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.  For the
step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item its improvement is about median in the benchmark suite
  considered in section~\ref{sec:eval}, and thus it serves as a good
  sanity check for each of the optimization techniques considered.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract
interpreter according to the AAM approach.  In the initial
abstraction, each state carries a store (what is called per-state
store variance). The space of stores is exponential in size; without
further abstraction, the analysis is exponential and thus cannot
analyze the example in a reasonable amount of time.  In
section~\ref{sec:baseline}, we perform a further abstraction by
widening the store.  The resulting analyzer sacrifices precision for
speed and is able to analyze the example in about 1 minute.  This step
is described by Van Horn and Might~\cite[\S
3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make even
small examples feasible.  We therefore take a widened interpreter as
the baseline for our evaluation.

Section~\ref{sec:opt} gives a series of simple abstractions and
implementation techniques that, in total, speed up the analysis by
nearly a factor of 500, dropping the analysis time to a fraction of a
second.  Figure~\ref{fig:churchtime} shows the step-wise improvement
of the analysis time for this example.

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are (abstract) machine states and edges denote machine
transitions between states.

The techniques we propose for optimizing analysis fall into the
following categories:
\begin{enumerate}
\item generate fewer states by avoiding the eager exploration of
  non-deterministic choices that will later collapse into a single
  join point.  We accomplish this by applying lazy evaluation
  techniques so that non-determinism is evaluated \emph{by need}.

\item generate fewer states by avoiding unnecessary, intermediate
  states of a computation.  We accomplish this by applying compilation
  techniques from functional languages to avoid interpretive overhead
  in the machine transition system.

\item generate states faster.  We accomplish this by better algorithm
  design in the fixed-point computation we use to generate state graphs.
\end{enumerate}
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for the
small motivating example in Earl, et
al.~\cite{dvanhorn:Earl2012Introspective}.  By generating
significantly fewer states at a significantly faster rate, we are able
to achieve large performance improvements in terms of both time and
space.

Section~\ref{sec:eval} describes the evaluation of each optimization
technique applied to an implementation supporting a more realistic set
of features, including mutation, first-class control, compound data, a
full numeric tower and many more forms of primitive data and
operations.
%
We evaluate this implementation against a set of benchmark programs
drawn from the literature.
%
For all benchmarks, the optimized analyzer outperforms the baseline
by at least a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for Earl et. al. program.  Gray states
  follow variable references, {\tt ev} states are black, and all others
  are white. Part (a) shows the result of the baseline analyzer.  It
  has long ``corridor'' transitions and ``diamond'' subgraphs that
  fan-out from nondeterminism and fan-in from joins.  Part (b) shows
  the result of performing nondeterminism lazily and thus avoids many
  of the diamond subgraphs.  Part (c) shows the result of abstract
  compilation that removes interpretive overhead in the form of
  intermediate states, thus minimizing the corridor transitions.  The
  end result is a more compact abstraction of the program that can be
  generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

In this section, we give a brief review of the AAM approach by
defining a sound analytic framework for a core higher-order functional
language: Landin's ISWIM~\cite{dvanhorn:Landin1966Next}.  In the
subsequent sections, we will explore optimizations for the analyzer in
this simplified setting, but scaling these techniques to realistic
languages is straightforward and has been done for the analyzer
evaluated in section~\ref{sec:eval}.

ISWIM is a family of programming languages parameterized by a set of
base values and operations.  To make things concrete, we consider a
member of the ISWIM family with integers, booleans, and a few
operations.
%
Figure~\ref{fig:syntax} defines the syntax of ISWIM.  It
includes variables, literals (either integers, booleans, or
operations), $\lambda$-expressions for defining procedures, procedure
applications, and conditionals.  Expressions carry a label, $\mlab$,
which is drawn from an unspecified set and denotes the source location
of the expression; labels are used to disambiguate distinct, but
syntactically identical pieces of syntax.  We omit the label
annotation in contexts where it is irrelevant.

@@

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains};
figure~\ref{fig:aam} defines the transition relation (unmentioned
components stay the same).  The evaluation of a program is defined as
its set of traces that arise from iterating the machine transition
relation. The $\traces$ function produces the set of all proofs of
reachability for any state $\mstate$ from the injection of program
$\mexp$ (from which one could extract a string of states). The machine
is a very slight variation on a standard abstract machine for ISWIM in
``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It can be
systematically derived from a definitional interpreter through a
continuation-passing style transformation and defunctionalization, or
from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,\footnote{
More generally, we can have stores map to any domain that forms
 a Galois connection with sets of values, enabling $\interpdelta$ to
produce elaborate abstractions of base values (e.g., interval or octagon
abstractions). We use sets of values for a simpler exposition.
} not
  single values,

\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
\end{align*}
These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
used to determinize the choice of fresh addresses. The $\sqcup$ on
stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
relation is non-deterministic in its choice of addresses, however it
must always choose a fresh address when allocating a continuation or
variable binding.  If we consider machine states equivalent up to
consistent renaming and fix an allocation scheme, this relation
defines a deterministic machine (the relation is really a function).

The interpretation of primitive operations is defined by setting
$\interpdelta$ as follows:
\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

@@

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are
equivalent to the notion of a stuttering bisimulation, but are more
amenable to mechanisation (and thus rigorous proof). They also only
require reasoning over one step of the reduction relation, which makes
the proof concise. The trick is in defining a well-founded ordering
that determines when the two semantics will match up again, what
Manolios calls the pair of functions $\erankt$ and $\erankl$ (but we
don't need the generality that requires $\erankl$ since the uncompiled
semantics doesn't stutter at all).

@@

\begin{lemma}
Stores of value stacks completely abstract stacks of stores.
\end{lemma}
This depends on some wellformedness conditions about the order of the
stacks. The store of value stacks can be translated to a stack of
stores by taking successive ``snapshots'' of the store at different
timestamps from the max timestamp it holds down to 0. Vice versa, we
replay the changes across adjacent stores in the stack.

@@

The intermediate data structures of the above transfer function can all be streamlined into globals that are destructively updated. In particular, there are 5 globals:

@@

The reduction relation would then instead of building store deltas,
update the global store. We would also not view it as a general
relation, but a function that adds all next states to $F$ if they have
not already been seen. At the end of iterating through $F$, $\mseentime$ is
updated with the new states at the next timestamp. There is no cross-step store poisoning since the lookup is restricted to the current step's time, which points to the same value throughout the step.

@@

We have implemented, optimized, and evaluated an analysis framework
supporting higher-order functions, state, first-class control,
compound data, and a large number of primitive kinds of data and
operations such as floating point, complex, and exact rational
arithmetic.  The analysis is evaluated against a suite of Scheme benchmarks
drawn from the literature.
%
For each benchmark, we collect analysis times, peak memory usage (as
determined by Racket's GC statistics), and the rate of
states-per-second explored by the analysis for each of the
optimizations discussed in section~\ref{sec:opt}, cumulatively
applied.  The analysis is stopped after consuming 30 minutes of time
or 1 gigabyte of space.  When presenting \emph{relative} numbers, we
use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster),
thus giving a conservative estimate of improvements.

All benchmarks are calculated as an average of 5 runs, done in
parallel (each isolated to a core), on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz
with 12Gb of memory.

Many benchmarks cause the baseline analyzer to take longer than 30
minutes or to consume more than 1 gigabyte of memory, at which point the
analysis is stopped.  This is the case for the largest benchmark
program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the
most optimized analyzer.  For those benchmarks that did complete on
the baseline, the optimized analyzer outperformed the baseline by a
factor of two to three orders of magnitude.
% safer to not give exact factors
% 475 to 4,382.

@@

\item {\bf nucleic}: a floating-point intensive application taken from
  molecular biology that has been used widely in benchmarking
  functional language
  implementations~\cite{dvanhorn:Hartel1996Benchmarking} and analyses
  (e.g.~\cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
  It is a constraint satisfaction algorithm used to determine the
  three-dimensional structure of nucleic acids.

\item {\bf matrix} tests whether a matrix is maximal among all
  matrices of the same dimension obtainable by simple reordering of
  rows and columns and negation of any subset of rows and columns.  It
  is written in continuation-passing style (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf nbody}: implementation~\cite{ianjohnson:nbody87} of the
  Greengard multipole algorithm for computing gravitational forces on
  point masses distributed uniformly in a cube (used in
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf earley}: Earley's parsing algorithm, applied to a 15-symbol
  input according to a simple ambiguous grammar.  A real program,
  applied to small data whose exponential behavior leads to a peak
  heap size of half a gigabyte or more during concrete execution.

@@

\item {\bf graphs}: counts the number of directed graphs with a
  distinguished root and \(k\) vertices, each having out-degree at
  most 2. It is written in a continuation-passing style and makes
  extensive use of higher-order procedures---it creates closures
  almost as often as it performs non-tail procedure calls (used by
  \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

Figure~\ref{fig:bench-overview} gives an overview of the benchmark
results in terms of absolute time, space, and speed between the
baseline and most optimized analyzer.  Figure~\ref{fig:bench-all}
plots the factors of improvement over the baseline for each
optimization step.

>>>>

\begin{abstract}
The technique of \emph{abstracting abstract machines} (AAM) provides a systematic approach for deriving computable approximations of evaluators that are easily proved sound.
%
This article contributes a complementary step-by-step process for subsequently going from a \naive{} analyzer derived under the AAM approach, to an efficient and correct implementation.
%
The end result of the process is a two to three order-of-magnitude improvement over the systematically derived analyzer, making it competitive with hand-optimized implementations that compute fundamentally less precise results.
\end{abstract}

@@

Program analysis provides sound predictive models of program behavior, but in order for such models to be effective, they must be efficiently computable and correct.
%
Past approaches to designing program analysis have often featured abstractions that are far removed from the original language semantics, requiring ingenuity in their construction and effort in their verification.
%
The \emph{abstracting abstract machines} (AAM) approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic} to deriving program analyses provides an alternative: a systematic way of transforming a programming language semantics in the form of an abstract machine into a family of abstract interpreters.
%
It thus reduces the burden of constructing and verifying the soundness of an abstract interpreter.
%

@@

By taking a machine-oriented view of computation, AAM makes it possible to design, verify, and implement program analyzers for realistic language features typically considered difficult to model.
%
The approach was originally applied to features such as higher-order functions, stack inspection, exceptions, laziness, first-class continuations, and garbage collection.
%
It has since been used to verify actor-\cite{local:DOsualdo:12A} and thread-based~\cite{dvanhorn:Might2011Family} parallelism and behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder};
%
it has been used to model Coq~\cite{local:harvard}, Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B}, JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.
%%

%%
The primary strength of the approach is that abstract interpreters can be easily derived through a small number of steps from existing machine models.
%
Since the relationships between abstract machines and higher-level semantic models---such as definitional interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from these high-level semantic models to sound program analyzers in a systematic way.
%
Moreover, since these analyses so closely resemble a language's interpreter
%
(a) implementing an analysis requires little more than implementing an interpreter,
%
(b) a single implementation can serve as both an interpreter and analyzer, and
%
(c) verifying the correctness of the implementation is straightforward.
%%

%%
Unfortunately, the AAM approach yields analyzers with poor performance relative to hand-optimized analyzers.
%
Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an equally systematic way through a number of simple steps, many of which are inspired by run-time implementation techniques such as laziness and compilation to avoid interpretative overhead.
%
Each of these steps is proven correct, so the end result is an implementation that is trustworthy and efficient.
%%

%%
In this article, we develop a systematic approach to deriving a practical implementation of an abstract-machine-based analyzer using mostly semantic means rather than tricky engineering.
%
Our goal is to empower programming language implementers and researchers to explore and convincingly exhibit their ideas with a low barrier to entry.
%
The optimizations we describe are widely applicable and apparently effective to scale far beyond the size of programs typically considered in the recent literature on flow analysis for functional languages.

@@

\begin{figure}[t]
\small
\begin{center}
%\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{
Factor improvements over the baseline analyzer for the \Church{} benchmark in terms of the rate of state transitions and total analysis time.
%
(Bigger is better.)
%
Each point is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

We start with a quick review of the AAM approach to develop an analysis framework and then apply our step-by-step optimization techniques in the simplified setting of a core functional language.
%
This allows us to explicate the optimizations with a minimal amount of inessential technical overhead.
%
Following that, we scale this approach up to an analyzer for a realistic untyped, higher-order imperative language with a number of interesting features and then measure improvements across a suite of benchmarks.
%%

%%
At each step during the initial presentation and development, we evaluated the implementation on a set of benchmarks.
%
The highlighted benchmark in figure \ref{fig:churchtime} is from Vardoulakis and Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity of multiplication over addition on Church numerals.
%
For the step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with complicated environment and control structure arising from the use of higher-order functions to encode arithmetic, and
%
\item its improvement is about median in the benchmark suite considered in section~\ref{sec:eval}, and thus it serves as a good sanity check for each of the optimization techniques considered.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract interpreter according to the AAM approach.
%
In the initial abstraction, each state carries a store (what is called per-state store variance).
%
The space of stores is exponential in size; without further abstraction, the analysis is exponential and thus cannot analyze the example in a reasonable amount of time.
%
In section~\ref{sec:baseline}, we perform a further abstraction by widening the store.
%
The resulting analyzer sacrifices precision for speed and is able to analyze the example in about 1 minute.
%
This step is described by Van Horn and Might~\cite[\S 3.5--6]{dvanhorn:VanHorn2012Systematic} and is necessary to make even small examples feasible.
%
We therefore take a widened interpreter as the baseline for our evaluation.
%%

%%
Section~\ref{sec:opt} gives a series of simple abstractions and implementation techniques that, in total, speed up the analysis by nearly a factor of 500, dropping the analysis time to a fraction of a second.
%
Figure~\ref{fig:churchtime} shows the step-wise improvement of the analysis time for this example.
%%

%%
The AAM approach, in essence, does the following: it takes a machine-based view of computation and turns it into a \emph{finitary approximation} by bounding the size of the store.
%
With a limited address space, the store must map addresses to \emph{sets} of values.
%
Store updates are interpreted as joins, and store dereferences are interpreted by non-deterministic choice of an element from a set.
%
The result of analyzing a program is a finite directed graph where nodes in the graph are (abstract) machine states and edges denote machine transitions between states.
%%

%%
The techniques we propose for optimizing analysis fall into the following categories:
\begin{enumerate}
\item generate fewer states by avoiding the eager exploration of non-deterministic choices that will later collapse into a single join point.
%
  We accomplish this by applying lazy evaluation techniques so that non-determinism is evaluated \emph{by need}.

\item generate fewer states by avoiding unnecessary, intermediate states of a computation.
%
  We accomplish this by applying compilation techniques from functional languages to avoid interpretive overhead in the machine transition system.

\item generate states faster.
%
  We accomplish this by better algorithm design in the fixed-point computation we use to generate state graphs.
\end{enumerate}
%
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for the small motivating example in Earl, et al.~\cite{dvanhorn:Earl2012Introspective}.
%
By generating significantly fewer states at a significantly faster rate, we are able to achieve large performance improvements in terms of both time and space.
%%

%%
Section~\ref{sec:eval} describes the evaluation of each optimization technique applied to an implementation supporting a more realistic set of features, including mutation, first-class control, compound data, a full numeric tower and many more forms of primitive data and operations.
%
We evaluate this implementation against a set of benchmark programs drawn from the literature.
%
For all benchmarks, the optimized analyzer outperforms the baseline by at least a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{
Example state graphs for Earl et. al. program.
%
  Gray states follow variable references, {\tt ev} states are black, and all others are white.
%
 Part (a) shows the result of the baseline analyzer.
%
  It has long ``corridor'' transitions and ``diamond'' subgraphs that fan-out from nondeterminism and fan-in from joins.
%
  Part (b) shows the result of performing nondeterminism lazily and thus avoids many of the diamond subgraphs.
%
  Part (c) shows the result of abstract compilation that removes interpretive overhead in the form of intermediate states, thus minimizing the corridor transitions.
%
  The end result is a more compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

In this section, we give a brief review of the AAM approach by
defining a sound analytic framework for a core higher-order functional
language: Landin's ISWIM~\cite{dvanhorn:Landin1966Next}.
%
In the subsequent sections, we will explore optimizations for the analyzer in this simplified setting, but scaling these techniques to realistic languages is straightforward and has been done for the analyzer evaluated in section~\ref{sec:eval}.
%%

%%
ISWIM is a family of programming languages parameterized by a set of base values and operations.
%
To make things concrete, we consider a member of the ISWIM family with integers, booleans, and a few operations.
%
Figure~\ref{fig:syntax} defines the syntax of ISWIM.
%
It includes variables, literals (either integers, booleans, or operations), $\lambda$-expressions for defining procedures, procedure applications, and conditionals.
%
Expressions carry a label, $\mlab$, which is drawn from an unspecified set and denotes the source location of the expression; labels are used to disambiguate distinct, but syntactically identical pieces of syntax.
%
We omit the label annotation in contexts where it is irrelevant.

@@

The semantics are defined in terms of a machine model.
%
The machine components are defined in figure~\ref{fig:domains};
%
figure~\ref{fig:aam} defines the transition relation (unmentioned components stay the same).
%
The evaluation of a program is defined as its set of traces that arise from iterating the machine transition relation.
%
The $\traces$ function produces the set of all proofs of reachability for any state $\mstate$ from the injection of program $\mexp$ (from which one could extract a string of states).
%
 The machine is a very slight variation on a standard abstract machine for ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.
%
It can be systematically derived from a definitional interpreter through a continuation-passing style transformation and defunctionalization, or from a structural operational semantics using the refocusing construction of Danvy and Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
%%

%%
Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,
%
\footnote{
More generally, we can have stores map to any domain that forms a Galois connection with sets of values, enabling $\interpdelta$ to produce elaborate abstractions of base values (e.g., interval or octagon abstractions).
%
We use sets of values for a simpler exposition.
}
%
not single values,

\item continuations are heap-allocated, not stack-allocated,
%
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax labels ($\mlab$) threaded through the computation, and
%
\item the machine is implicitly parameterized by the functions $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
\end{align*}
These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
used to determinize the choice of fresh addresses.
%
The $\sqcup$ on stores in the figure is a point-wise lifting of $\cup$:
%
 $\msto \sqcup \msto' = \lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$.
%
The resulting relation is non-deterministic in its choice of addresses, however it must always choose a fresh address when allocating a continuation or variable binding.
%
If we consider machine states equivalent up to consistent renaming and fix an allocation scheme, this relation defines a deterministic machine (the relation is really a function).
%%

%%
The interpretation of primitive operations is defined by setting $\interpdelta$ as follows:
\begin{align*}
\mnum+1 &\in \interpdelta(\saddone,\mnum) &
\mnum-1 &\in \interpdelta(\ssubone,\mnum)\\
\strue &\in \interpdelta(\szerohuh,\szero) &
\sfalse &\in \interpdelta(\szerohuh,\mnum)\text{ if }\mnum\neq \szero\\
\end{align*}

@@

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are
equivalent to the notion of a stuttering bisimulation, but are more
amenable to mechanization (and thus rigorous proof). They also only
require reasoning over one step of the reduction relation, which makes
the proof concise. The trick is in defining a well-founded ordering
that determines when the two semantics will match up again, what
Manolios calls the pair of functions $\erankt$ and $\erankl$ (but we
don't need the generality that requires $\erankl$ since the uncompiled
semantics doesn't stutter at all).

@@

\begin{lemma}
Stores of value stacks completely abstract stacks of stores.
\end{lemma}
This depends on some well-formedness conditions about the order of the
stacks. The store of value stacks can be translated to a stack of
stores by taking successive ``snapshots'' of the store at different
timestamps from the max timestamp it holds down to 0. Vice versa, we
replay the changes across adjacent stores in the stack.

@@

The intermediate data structures of the above transfer function can all be streamlined into globals that are destructively updated.
%
In particular, there are 5 globals:

@@

The reduction relation would then instead of building store deltas, update the global store.
%
We would also not view it as a general relation, but a function that adds all next states to $F$ if they have
not already been seen.
%
At the end of iterating through $F$, $\mseentime$ is updated with the new states at the next timestamp.
%
There is no cross-step store poisoning since the lookup is restricted to the current step's time, which points to the same value throughout the step.

@@

We have implemented, optimized, and evaluated an analysis framework supporting higher-order functions, state, first-class control, compound data, and a large number of primitive kinds of data and operations such as floating point, complex, and exact rational arithmetic.
%
The analysis is evaluated against a suite of Scheme benchmarks drawn from the literature.
%
For each benchmark, we collect analysis times, peak memory usage (as determined by Racket's GC statistics), and the rate of states-per-second explored by the analysis for each of the optimizations discussed in section~\ref{sec:opt}, cumulatively applied.
%
The analysis is stopped after consuming 30 minutes of time or 1 gigabyte of space.
%
When presenting \emph{relative} numbers, we use the timeout limits as a lower bound on the actual time required (i.e., one minute versus timeout is at least 30 times faster),
thus giving a conservative estimate of improvements.
%%

%%
All benchmarks are calculated as an average of 5 runs, done in
parallel (each isolated to a core), on an 12-core, 64-bit Intel Xeon machine running at 2.40GHz
with 12Gb of memory.
%%

%%
Many benchmarks cause the baseline analyzer to take longer than 30 minutes or to consume more than 1 gigabyte of memory, at which point the analysis is stopped.
%
This is the case for the largest benchmark program, {\bf nucleic}, which is 3,500 lines of code and takes under a minute in the most optimized analyzer.
%
For those benchmarks that did complete on the baseline, the optimized analyzer outperformed the baseline by a factor of two to three orders of magnitude.
% safer to not give exact factors
% 475 to 4,382.

@@

\item {\bf nucleic}: a floating-point intensive application taken from molecular biology that has been used widely in benchmarking functional language implementations~\cite{dvanhorn:Hartel1996Benchmarking} and analyses (e.g.~\cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
%
  It is a constraint satisfaction algorithm used to determine the three-dimensional structure of nucleic acids.

\item {\bf matrix} tests whether a matrix is maximal among all matrices of the same dimension obtainable by simple reordering of rows and columns and negation of any subset of rows and columns.
%
  It is written in continuation-passing style (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).

\item {\bf nbody}: implementation~\cite{ianjohnson:nbody87} of the Greengard multipole algorithm for computing gravitational forces on point masses distributed uniformly in a cube (used in \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
%%

%%
\item {\bf earley}: Earley's parsing algorithm, applied to a 15-symbol input according to a simple ambiguous grammar.
%
  A real program, applied to small data whose exponential behavior leads to a peak heap size of half a gigabyte or more during concrete execution.

@@

\item {\bf graphs}: counts the number of directed graphs with a distinguished root and \(k\) vertices, each having out-degree at most 2.
%
 It is written in a continuation-passing style and makes extensive use of higher-order procedures---it creates closures almost as often as it performs non-tail procedure calls (used by \cite{dvanhorn:wright-jagannathan-toplas98,dvanhorn:jagannathan-etal-popl98}).
\end{enumerate}

@@

Figure~\ref{fig:bench-overview} gives an overview of the benchmark results in terms of absolute time, space, and speed between the baseline and most optimized analyzer.
%
Figure~\ref{fig:bench-all} plots the factors of improvement over the baseline for each optimization step.
